---
title: "Gender Stereotypes in Early Development"
description: Exploring gender stereotypes among young children using permutation tests.
author: Maren Rusk
date: November 8, 2024
format: html
execute:
  warning: false
  message: false
---

```{r, echo=FALSE}

library(tidyverse)
library(openintro)
```

In this analysis, I will conduct a permutation test to attempt to answer the research question: Are young females less likely than young males to identify adults of their own gender as smart? The data for this test, [Gender Stereotypes in 5-7 year old Children]{.underline}, was found in the R openintro package. It reports the results of an experimental study conducted in 2017 by Lin Bian that tested whether some children were more likely to choose someone of their own gender when asked to identify a "smart" or "nice" individual in an image. The *stereotype* variable gives the proportion of trials where each child chose a gender target that matched their own gender. Focusing specifically on trials where the children were shown images of adults and asked to point out the person who appeared to be smart, I will test for a difference in this variable between the male and female genders. Below, I show a bar plot of the differences in stereotype proportion between male and female children.


```{r, fig.alt="This bar plot displays stereotype proportion by gender, as two side by side segmented bars, one male and one female. The bars are split up by the total number of each stereotype proportion observed, between the five possible proportions: 0, 0.25, 0.5, 0.75, and 1. There is a visible difference between the female and male bars, with males having a much larger count for larger stereotypes, and females having a larger count for smaller stereotypes."}

gender_stereo <- children_gender_stereo[["2"]] |>
  filter(trait == "smart", target == "adults")

head(gender_stereo)

prop_data <- gender_stereo |>
  mutate(stereotype = factor(stereotype))

ggplot(prop_data, aes(x = gender, fill = stereotype)) +
  geom_bar() +
  labs(
    title = "Stereotype proportion by gender for 'smart adult' selections"
  )
  
```
Based on this plot, we can see a clear difference between male and female stereotype proportions resulting from this study. In particular, the male bar shows a much larger count for an observed stereotype of 1, and the female bar shows a much larger count for an observed stereotype of 0.25. This means that 5-7 year old male children in the study were overall much more likely to choose male adults to be smart than female children were to choose female adults to be smart. The code chunk below shows that the difference in mean stereotype proportion between male and female children is in fact 0.0972.



```{r}
# wrangle data fo get difference in means between male/female
stereo_obs <- gender_stereo |>
  filter(trait == "smart", target == "adults") |>
  group_by(gender) |>
  summarize(mean_stereotype = mean(stereotype)) |>
  summarize(difference = diff(mean_stereotype)) |>
  pull(difference)

stereo_obs

```


Next, I shuffle the data and create a null sampling distribution, in order to determine whether this dataset contains statistically significant evidence of a stereotype difference.

```{r, fig.alt="This histogram displays a null sampling distribution of differences between stereotype proportion by gender using randomized data. The x-axis represents the mean difference in proportion of stereotype by gender, and the y-axis represents count. The shape of the histogram resembles a normal curve, with a slight gap at x = 0. A red vertical line intercepts the x-axis at x = 0.0972, the observed mean difference from the actual experimental data."}

# function to produce one permuted experiment under the null hypothesis
permute_stereo <- function(rep, data) {
  data |>
    select(gender, stereotype) |>
    mutate(stereo_perm = sample(stereotype, replace = FALSE)) |>
    group_by(gender) |>
    summarize(obs_avg = mean(stereotype), perm_avg = mean(stereo_perm)) |>
    summarize(obs_avg_diff = diff(obs_avg), perm_avg_diff = diff(perm_avg))
}

# simulate many experiments
set.seed(47)
num_exper <- 5000
null_dist <- map(1:num_exper, permute_stereo, data = gender_stereo) |>
  list_rbind()


# visualize null sampling distribution with p-value
null_dist |>
  ggplot(aes(x = perm_avg_diff)) +
  geom_histogram() +
  geom_vline(aes(xintercept = obs_avg_diff), color = "red") +
  theme_bw() +
  labs(
    title = "Sampling Distribution when Ho is True",
    subtitle = "aka no difference between gender picks",
    x = "mean difference in proportion of stereotype by gender"
  )

# calculating the p-value
null_dist |> 
  summarize(p_value = mean(perm_avg_diff > obs_avg_diff))
```

This plot displays a histogram of sample difference in means calculated under the assumption that there is no inherent difference between how often a male or female will pick their own gender to be smarter. The red line represents the observed mean difference stereotype proportion between the male and female genders from the data, 0.0972. The histogram allows for visualization of the above calculated p-value of 0.0302, meaning that only 3.02% of the 5000 samples had a difference in means (male - female) greater than 0.0972.

[Process]{.underline}

I began this test with the null hypothesis, Ho, that the difference between male and female stereotype was 0, and the alternative hypothesis, Ha, that the difference between male and female stereotype was greater than 0:

Ho: μmale - μfemale = 0, Ha: μmale - μfemale \> 0

I wrangled my dataset to find the mean stereotype proportion for all females and all males in the study, and ultimately the difference in these means (0.0972). Then, I created a null sampling distribution under the assumption of the null hypothesis by shuffling the experimental data 5000 times to randomize the pairing between gender and stereotype proportions, and finding the difference in male and female means for each experiment. I used this distribution to calculate a p-value by finding the percent of experiments where the difference means (male - female) was greater than the observed difference in means from the dataset.

[Results]{.underline}

I found a p-value of 0.0302, which is low enough to reject the null hypothesis. Thus, this dataset shows statistically significant evidence that 5-7 year old males are more likely to select adults of their own gender to be smart than 5-7 year old females are.

[Data Source]{.underline}

Bian L, Leslie SJ, Cimpian A. 2017. "Gender stereotypes about intellectual ability emerge early and influence children's interests". Science 355:6323 (389-391). https://www.science.org/doi/10.1126/science.aah6524.
