---
title: "Text Analysis"
description: Analysis of New York Times Headlines Data from RTextTools Package
author: Maren Rusk
date: October 2, 2024
format: html
execute:
  warning: false
  message: false
---

Question 1: Which president (Bill Clinton or George W. Bush) was mentioned in more NYT Headlines during their respective presidential terms (Clinton 1997-2000 / Bush 2001-2004)?


```{r}
#| echo: false
library(RTextTools) 
library(tidyverse)
data(NYTimes)

```



```{r}
NYTimes <- NYTimes |>
 
  # create a column for year using date
  mutate(Year_str = str_extract(Date, "\\D-\\d\\d")) |>
  mutate(Year = str_sub(Year_str, 3, 4))

# Headlines with "Clinton" between 1997 and 2000  
Clinton_Headlines <- NYTimes |>
  filter(Year >= 97 | Year == "00") |>
  filter(str_detect(Title, "Clinton")) |>
  select(Date, Year, Title) |>
  mutate(President = "Clinton")

# Headlines with "Bush" between 2001 and 2004
Bush_Headlines <- NYTimes |>
  filter(Year <= 04, Year != "00", Year != "05", Year != "06") |>
  filter(str_detect(Title, "Bush")) |>
  select(Date, Year, Title) |>
  mutate(President = "Bush")

#Join these two together to plot
Both_Headlines <- full_join(Clinton_Headlines, Bush_Headlines) |>
  group_by(President, Year) |>
  summarize(num_headlines = n())

# bar chart
ggplot(Both_Headlines, aes(x = fct_rev(President), y = num_headlines, 
                           fill = fct_relevel(Year, c("97", "98","99", "00", "01", "02", "03", "04")))) + 
  geom_col() + 
  scale_x_discrete(labels=c('Clinton (1997-2000)', 'Bush (2001-2004)')) + 
  labs(
    title = "Clinton v Bush: Which President Had More Headlines During Their Term?",
    x = "President",
    y = "Number of Mentions in NYT Headlines During Term",
    fill = "Year"
  ) + 
  theme_classic()
  
```
This plot shows the number of New York Times headlines that each president was mentioned in during their consecutive terms (Bill Clinton 1997-2000 and George W. Bush 2001-2004) by year. Based on the plot, there is clearly a large difference in both the total number of headlines for each president, where Bush had a total of 65 headline mentions between 2001 and 2004 and Clinton only had 20. There does not seem to be a consistent pattern of increase or decrease of headlines by year during a presidential term. Notably, Bush had a huge surge of headlines in 2001 which is likely due to the influx of New York Times articles around the crisis of 9/11. It would be interesting to look at whether there was also a large difference in the total number of New York Times articles in this data set between 1997-2000 and 2001-2004 that might confound these results.



Question 2: Does the typical number of words per headline vary by month? Does this relationship change depending on
whether the headline contains punctuation?

```{r}

NYTimesWM <- NYTimes |>
  # extract month from date as new variable
  mutate(Month = str_extract(Date, "(?<=-).+(?=-)")) |>
  mutate(word_list = str_split(Title, " ")) |>
  mutate(num_words = lengths(word_list)) |>
  # variable for whether headline contains punctuation
  mutate(Punctuation = grepl("[[:punct:]]", Title))


# boxplot  
ggplot(NYTimesWM, aes(x = fct_relevel(Month, c('Jan', 'Feb', 'Mar',
                                               'Apr', 'May', 'Jun',
                                               'Jul', 'Aug', 'Sep',
                                               'Oct', 'Nov', 'Dec')), 
                      y = num_words, color = Punctuation)) + 
  geom_boxplot() + 
  labs(
    title = "Length of New York Times Headlines by Month 
    With and Without Punctuation Between 1996 and 2006",
    x = "Month",
    y = "Number of Words Per Headline"
  )

```
This plot shows the distributions of the number of words per headline for each month for headlines with and without punctuation. By month, the distributions do not show a clear trend, but the ranges for February and November with punctuation appear to be the largest overall excluding outliers, while May has both the longest headline (largest outlier) and one of the shortest, at one word. However, looking at the comparison between headlines with punctuation and headlines without punctuation, the ranges for headlines with punctuation are longer for every single month, and there are also far more outliers on the high end for this category. In fact, the upper quartile for headlines with punctuation is a larger number of words for every month except August, which is even with no punctuation. These results are understandable considering that longer sentences are more likely to need punctuation in order to have proper grammar. The months with the largest difference between punctuation and no punctuation in median number of words per headline are September and November.

