---
title: "Text Analysis"
description: My analysis
author: Maren Rusk
date: September 18, 2024
format: html
execute:
  warning: false
  message: false
---

Here is my text analysis of NYT Headlines!

Questions of Interest:
1. Which president (Bill Clinton or George W. Bush) were mentioned in more NYT Headlines during their respective presidential terms (Clinton 1997-2000 / Bush 2001-2004)?

- large difference
- is this because it was clinton's 2nd term/ bush's first term? or because more happened during bush's term (9/11)? or was there more overall NYT coverage then?

2. Have headlines about the supreme court become more frequent over time (between 1996 and 2006)?

3. Does the average length/number of the words in NYT headlines vary by month? year?



```{r}
library(RTextTools) 
library(tidyverse)
data(NYTimes)
as_tibble(NYTimes)

```



```{r}

NYTimes <- NYTimes |>
  # filter by date (only Jan 1 97 - Dec 31 2004)
  
  mutate(Year_str = str_sub(str_extract(Date, "\\D-\\d\\d")), 3, 4) |>
  mutate(Year = str_sub(Year_str, 3, 4))
  
Clinton_Headlines <- NYTimes |>
  filter(Year >= 97 | Year == "00") |>
  filter(str_detect(Title, "Clinton")) |>
  select(Date, Year, Title) |>
  mutate(President = "Clinton")

Clinton_Headlines

Bush_Headlines <- NYTimes |>
  filter(Year <= 04, Year != "00", Year != "05", Year != "06") |>
  filter(str_detect(Title, "Bush")) |>
  select(Date, Year, Title) |>
  mutate(President = "Bush")

Bush_Headlines

Both_Headlines <- full_join(Clinton_Headlines, Bush_Headlines) |>
  group_by(President, Year) |>
  summarize(num_headlines = n())
Both_Headlines

#Join these two together to plot

ggplot(Both_Headlines, aes(x = President, y = num_headlines, color = President)) + 
  geom_col() + 
  labs(
    title = "Which President Had More Headlines During Their Term?",
    y = "Number of Mentions in NYT Headlines During Term"
  ) + 
  theme_classic()
  
```

