---
title: "Olympics"
description: Analysis of Olympic athletes from TidyTuesday
author: Maren Rusk
date: September 18, 2024
format: html
execute:
  warning: false
  message: false
---

**Question:** How does the average age of Olympic participants vary by sport and how has that changed over time? *(I have selected 6 sports from the data for readability purposes.)*

**Code:**

```{r}

knitr::opts_chunk$set(error = TRUE)

library(tidyverse)
library(ggplot2)
library(dplyr)

olympics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-08-06/olympics.csv')
```

```{r}
selected_sports <- filter(olympics, sport == "Swimming" | sport == "Gymnastics" | sport == "Equestrianism" | sport == "Diving" | sport == "Weightlifting" | sport == "Speed Skating")

new_olympics <- selected_sports |>
  group_by(year, sport) |>
  mutate(avg_age = mean(age, na.rm = TRUE))

ggplot(new_olympics, aes(x = year, y = avg_age, color = sport)) + 
  geom_line() + 
  labs(
    title = "Average Age of Olympic Participants by Sport",
    x = "Olympic Year", 
    y = "Average age of Participants"
  )



```

**Analysis:**

From this plot, we can see that the average age of Equestrian Olympians has been consistently 5+ years greater than the averages of the other five sports shown between 1890 and 2010. Notably, where most sports saw a large increase in average age between the late 1800s and early 1900s, equestrianism shows a large decrease in average participant age during these years.

Among the other sports, swimming consistently had the lowest average age of competitors, until Gymnastics surpassed it in the mid-1980s. Meanwhile, weightlifting has average ages of its Olympians most consistently on the high end, but still nowhere near the equestrian average.

Overall, there does not appear to be a very strong relationship between year and average age of Olympic participants that is consistent across these sports, except an all-around dip in age around 1980.

**Link to Data:** <https://github.com/rfordatascience/tidytuesday/tree/master/data/2024/2024-08-06>
