---
title: "Olympics Exploration"
description: Analysis of Olympic athletes from TidyTuesday
author: Maren Rusk
date: September 18, 2024
format: html
execute:
  warning: false
  message: false
---

In this project, I will use data from the "Olympic Athletes and Medals" folder from the TidyTuesday github repository. This data was originally scraped from <https://www.sports-reference.com/> into a Kaggle dataset by RGriffin, and contains bio and medal results information on athletes from the 1896 Athens games through Rio 2016.

I have selected six sports from the dataset: swimming, gymnastics, equestrianism, diving, weightlifting, and speed skating, with the intent to explore how the average age of Olympic participants varies by sport and how that has changed over time, using visuals.

An example of the data can be seen below.

```{r, echo=FALSE}

knitr::opts_chunk$set(error = TRUE)
library(tidyverse)
library(ggplot2)

olympics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-08-06/olympics.csv') 
```

```{r}
olympics <- olympics |>
  filter(sport %in% c("Swimming", "Gymnastics", "Equestrianism", "Diving", "Weightlifting", "Speed Skating")) |>
  select(name, year, sport, team, age, event, medal)

head(olympics, 10)
```

Then, I wrangled the data to find the average age of athletes from each year and selected sport. Note that not all of these sports were part of the Olympics in 1896, so average values only appear for the years that the sport was contested in the Olympics.

```{r}
# calculate average age in each sport by olympic year, to be plotted
olympics_plot <- olympics |>
  group_by(year, sport) |>
  summarize(avg_age = mean(age, na.rm = TRUE))

head(olympics_plot, 10)

```

Next, I created a plot of this data of year vs average age, separating each sport into a different colored line.

```{r, fig.alt= "This plot shows the average age of olympic participants in each sport from the years 1896 to 2016. The x-axis represents the year for each Olympics, and the y-axis represents the average age of participants in a given sport for that Olympic games. The data points are marked as colored points along a separate trend line for each of the six sports. The lines follow similar trajectories, each showing a bit of a peak around 1950 before falling, and trending slowly back upwards in average age approaching current day. The average age for equestrian olympians is about 10-15 years higher than each of the other sports, which are more clumped together between 18 and 28 years of age."}

ggplot(olympics_plot, aes(x = year, y = avg_age, color = sport)) + 
  geom_point() + 
  geom_line() +
  labs(
    title = "Average Age of Olympic Participants by Sport",
    x = "Olympic Year", 
    y = "Average age of Participants"
  ) +
  theme_bw()

```

**Analysis:**

From this plot, we can see that the average age of Equestrian Olympians has been consistently 5+ years greater than the averages of the other five sports shown between 1890 and 2010. Notably, where most sports saw a large increase in average age between the late 1800s and early 1900s, equestrianism shows a large decrease in average participant age during these years.

Among the other sports, swimming consistently had the lowest average age of competitors, until Gymnastics surpassed it in the mid-1980s. Meanwhile, weightlifting has average ages of its Olympians most consistently on the high end, but still nowhere near the equestrian average.

Overall, there does not appear to be a very strong relationship between year and average age of Olympic participants that is consistent across these sports, except an all-around dip in age around 1980.

**Link to Data:** <https://github.com/rfordatascience/tidytuesday/tree/master/data/2024/2024-08-06>
